Efficient Handling of String-Number Conversion

### 2020-03-30 final-version/evaluation.tex
small fixes related to rev C and few other minor things

%more or less standard
The way our theory solver and Z3 interact is almost standard. When Z3 asks our theory solver a string constraint satisfiability problem, our solver tries to prove it is SAT or UNSAT using the procedure discuss in this paper. For under-approximation, whenever a corresponding linear formula is created, we attach the current value of $m$, $p$, $q$ to the formula, and then push it to Z3 core. If our theory solver reports UNKNOWN, Z3 remembers it in a global flag \textsf{incomplete} and either tries another solution branch, or the same solution branch with different value of $m$, $p$, $q$. If Z3 completes the search of all solution branches, it reports UNSAT if the flag \textsf{incomplete} is down, and UNKNOWN otherwise.

>>>>

%more or less standard
The way our theory solver and Z3 interact is almost standard. When Z3 asks our theory solver a string constraint satisfiability problem, our solver tries to prove it is SAT or UNSAT using the procedure discussed in this paper. For under-approximation, whenever a corresponding linear formula is created, we attach the current value of $m$, $p$, $q$ to the formula, and then push it to Z3 core. If our theory solver reports UNKNOWN, Z3 remembers it in a global flag \textsf{incomplete} and either tries another solution branch, or the same solution branch with different value of $m$, $p$, $q$. If Z3 completes the search of all solution branches, it reports UNSAT if the flag \textsf{incomplete} is down, and UNKNOWN otherwise.

### 2020-03-30 final-version/main.tex
small fixes related to rev C and few other minor things

In the following, we explain the construction of the linear formula using $\Phi$ as an example. Assume that we project the domains of $x$ and $y$ to the PFA in Figure~\ref{fig:sfa} (a) and (b), respectively. The variables $v_1$, $v_2$, $v_3$, $v_4$ in the figure are \emph{character} variables. Thus, $v_1$, $v_2$, $v_3$, $v_4$ are also integer variables. For example, from  $ y \in (12)^+$, we may derive $v_2=1$ and $v_3=2$.

The linear formula produced after the domain restriction will be over variables $v_1$, $v_2$, $v_3$, $v_4$, as well as the number of occurrences of each character variable $\#v_1$, $\#v_2$, $\#v_3$, $\#v_4$. Each model of the linear formula encodes a model of the string constraint. For example, $x=``12"$ and $y=``1212"$ is encoded by the assignment $(v_1,v_2,v_3,v_4,\#v_1,\#v_2,\#v_3,\#v_4) \rightarrow (1,2,1,2,1,1,2,2)$.% 
\footnote{In these examples, we use the shorthand $(x_1,\ldots,x_k)\rightarrow(n_1,\ldots,n_k)$ to denote the function $\{x_1 \mapsto n_1,\ldots, x_k\mapsto n_k\}$.}
The assignment says, for example, that $x$ is the \emph{parametric word} obtained by traversing the loop of $A_x$ once (because $\#v_1 = \#v_2 = 1$), which is $v_1v_2$. Under the assignment $v_1=1$ and $v_2=2$, we obtain $x=``12"$.

@@

The \emph{Parikh image} of a word $w\in \Sigma^*$ maps each \emph{Parikh variable} $\#a$, with $a \in \Sigma$ is a character, to the number of occurrences of $a$ in $w$.
Formally, given a set $S$, let $\pvarsof S$ denote the set of Parikh variables $\{\#s \mid s\in S\}$. 
The Parikh image of $w$ is a function $\parikhof w:\pvarsof \Sigma \rightarrow \nat$ such that $\parikhof w (\#a) = |w|_a$. 
The Parikh image of a language $L$ is defined as follows $\parikhof L = \{\parikhof w \mid w \in L\}$. %and the Parikh image of an FA $A$ is Parikh image of its language, $\parikhof{A} = \parikhof{L(A)}$. 
%
It is well known that the Parikh image of a regular language  can be  characterized by a linear formula:

@@

Assume that $\vars$ ia a set of \emph{string variables} ranging over $\Sigma^*$ and $\varn$ a set of \emph{integer variables} ranging over $\mathbb{Z}$.
An \emph{interpretation over $\vars$ and $\varn$} is a mapping $I:\vars\cup \varn\rightarrow\Sigma^* \cup \mathbb{Z}$. A \emph{word term} is an element in $\vars^*$. We lift the interpretation $I$ to word terms and linear constraints in the standard manner.

@@

\item A \emph{string-number conversion constraint} $\phi_s$ is of the form $n=\sti{x}$, where the function $\sti{x}$ is defined as follows. For $a\in [0,9]$, we have $\sti{a}=a$ and for $w \cdot a \in [0,9]^+$, $\sti{w\cdot a} = 10\times \sti{w}+a$. For $w\notin [0,9]^+$, $\sti{w}=-1$. We define $\modelsof{\phi_s}= \{I \mid I(n)= \sti{I(x)} \}$. The \emph{number-string} conversion constraint $x=\its{n}$ is treated as a syntactic sugar for $n=\sti{x}$.

@@

\begin{lemma}\label{lemma:decoding}
For an flat FA $A$, there is a function $\decode A$ such that for each $w\in L(A)$, $\decode A(\parikhof w) = w$. 
\end{lemma}
%
Observe that the $\parikhof w$ value of any variable appearing within  a cycle of $A$ is equal to the number of repetitions of that cycle in the accepting run. This is an immediate consequence of the fact that 
 every character appears on at most one transition. Thus, the accepting run on $w$ (and so $w$ itself) can be reconstructed from $\parikhof w$.

@@

Consequently, the automaton $A^m$ of our number PFA will have the following form illustrated in Figure~\ref{fig:sfa_its}. 
It has a self-loop on the initial state labeled by the character variable $v_0$, 
forced by the constraint 
$$
\Psi_{v_0}\defeq v_0=0
$$ 
to hold the value $0$. 
This transition ensures that the under-approximation handles numerals with arbitrary number of trailing zeros.  
The self-loop is followed by a chain of $m$ transitions $(q_{i-1},v_i,q_{i}),1\leq i \leq m$, leading towards the final state $q_m$. 
The chain encodes at most $m$ meaningful digits (only \emph{at most} because the first variables in the sequence can still be assigned zeros and some variables may be assigned $\epsilon$). Hence this PFA covers all numerals that encode numbers with at most $m$ digits.
Although it still has a loop, it will not create any exponential component defining the value of $n$ because the loop only represents a sequence of ``0" at the front of $x$. Thus, it will not affect the integer value of $n=\sti{x}$.

@@

Numeric PFA with these restrictions would satisfy our primary objective, that is, they would induce linear formulae and would ``easily" and completely cover all numerals.
%
A last problem still needs to be solved before they can be efficient in practice. 
%
Recall that the character variables can be assigned $\epsilon$.  
%
Therefore, a single chain of $k$ interesting digits, $k\leq m$, can be by $A^m$ represented in $k\choose m$ ways, each corresponding to one possible interleaving of $k$ digits with $m-k$ epsilons. 
%
%\lh{I am not sure whether it is really impossible to make a small formula}
%
This may  lead to a formula of   exponential size  when defining  the value of $n$.
%
In order to eliminate this potential blow-up in the size of the formula, we add to $\paf^m$ an additional constraint that forces all epsilons to be \emph{shifted} behind the least significant digit. This will leave us with only one interleaving. This is the formula
$$
\Psi_{\mathsf{shift}}^m \defeq \bigwedge_{1\leq i \leq m} v_i \neq \noepsilon \implies v_{i-1}\neq \noepsilon \ .
$$
Last, since this restriction is meaningful only when the string is indeed a numeral,  
we also define the constraint representing the strings which are not numerals, the formula
$$
\Psi_{\mathsf{NaN}}^m \defeq  \bigvee_{i\in [1,m]} v_i>9
$$ 
and define the final form of the interpretation restriction used by $A^m$ as 
$$
\paf^m \defeq  \Psi_{\mathsf{NaN}}^m \lor (\Psi_{v_0} \land \Psi_{\mathsf{shift}}^m)\ .
$$
Consequently, we design our domain restrictions $\dr$ so that for string variables $x$ that appear within string-integer constraints,  $\dr(x)$ is a number PFA $(A^m,\paf^m),m\in \nat$.

>>>>

In the following, we explain the construction of the linear formula using $\Phi$ as an example. Assume that we project the domains of $x$ and $y$ to the PFA in Figure~\ref{fig:sfa} (a) and (b), respectively. The variables $v_0$, $v_1$, $v_2$, $v_3$ in the figure are \emph{character} variables. Thus, $v_0$, $v_1$, $v_2$, $v_3$ are also integer variables. For example, from  $ y \in (12)^+$, we may derive $v_1=1$ and $v_2=2$.

The linear formula produced after the domain restriction will be over variables $v_0$, $v_1$, $v_2$, $v_3$, as well as the number of occurrences of each character variable $\#v_0$, $\#v_1$, $\#v_2$, $\#v_3$. Each model of the linear formula encodes a model of the string constraint. For example, $x=``12"$ and $y=``1212"$ is encoded by the assignment $(v_0,v_1,v_2,v_3,\#v_0,\#v_1,\#v_2,\#v_3) \rightarrow (1,2,1,2,1,1,2,2)$.% 
\footnote{In these examples, we use the shorthand $(x_1,\ldots,x_k)\rightarrow(n_1,\ldots,n_k)$ to denote the function $\{x_1 \mapsto n_1,\ldots, x_k\mapsto n_k\}$.}
The assignment says, for example, that $x$ is the \emph{parametric word} obtained by traversing the loop of $A_x$ once (because $\#v_0 = \#v_1 = 1$), which is $v_0v_1$. Under the assignment $v_0=1$ and $v_1=2$, we obtain $x=``12"$.

@@

The \emph{Parikh image} of a word $w\in \Sigma^*$ maps each \emph{Parikh variable} $\#a$, where $a \in \Sigma$ is a character, to the number of occurrences of $a$ in $w$.
Formally, given a set $S$, let $\pvarsof S$ denote the set of Parikh variables $\{\#s \mid s\in S\}$. 
The Parikh image of $w$ is a function $\parikhof w:\pvarsof \Sigma \rightarrow \nat$ such that $\parikhof w (\#a) = |w|_a$ for each $a\in\Sigma$.
The Parikh image of a language $L$ is defined as follows $\parikhof L = \{\parikhof w \mid w \in L\}$. %and the Parikh image of an FA $A$ is Parikh image of its language, $\parikhof{A} = \parikhof{L(A)}$. 
%
It is well known that the Parikh image of a regular language  can be  characterized by a linear formula:

@@

Assume that $\vars$ is a set of \emph{string variables} ranging over $\Sigma^*$ and $\varn$ a set of \emph{integer variables} ranging over $\mathbb{Z}$.
An \emph{interpretation over $\vars$ and $\varn$} is a mapping $I:\vars\cup \varn\rightarrow\Sigma^* \cup \mathbb{Z}$. A \emph{word term} is an element in $\vars^*$. We lift the interpretation $I$ to word terms and linear constraints in the standard manner.

@@

\item A \emph{string-number conversion constraint} $\phi_s$ is of the form $n=\sti{x}$, where the function $\sti{x}$ is defined as follows. For $a\in [0,9]$, we have $\sti{a}=a$ and for $w \cdot a \in [0,9]^+$, $\sti{w\cdot a} = 10\times \sti{w}+a$. For $w\notin [0,9]^+$, $\sti{w}=-1$. We define $\modelsof{\phi_s}= \{I \mid I(n)= \sti{I(x)} \}$. The \emph{number-string} conversion constraint $x=\its{n}$ is treated as a syntactic sugar for $n=\sti{x}$. We assume decimal encoding of numbers.

@@

\begin{lemma}\label{lemma:decoding}
For a flat FA $A$, there is a function $\decode A$ such that for each $w\in L(A)$, $\decode A(\parikhof w) = w$. 
\end{lemma}
%
Observe that the $\parikhof w$ value of any variable appearing within  a cycle of $A$ is equal to the number of repetitions of that cycle in the accepting run. This is an immediate consequence of the fact that 
 every character appears on at most one transition. Thus, the accepting run on $w$ (and so $w$ itself) can be reconstructed from $\parikhof w$.

@@

Consequently, the automaton $A^m$ of our numeric PFA will have the following form illustrated in Figure~\ref{fig:sfa_its}. 
It has a self-loop on the initial state labeled by the character variable $v_0$, 
forced by the constraint 
$$
\Psi_{v_0}\defeq v_0=0
$$ 
to hold the value $0$. 
This transition ensures that the under-approximation handles numerals with arbitrary number of trailing zeros.  
The self-loop is followed by a chain of $m$ transitions $(q_{i-1},v_i,q_{i}),1\leq i \leq m$, leading towards the final state $q_m$. 
The chain encodes at most $m$ meaningful digits (only \emph{at most} because the first variables in the sequence can still be assigned zeros and some variables may be assigned $\epsilon$). Hence this PFA covers all numerals that encode numbers with at most $m$ digits.
Although it still has a loop, it will not create any exponential component defining the value of $n$ because the loop only represents a sequence of ``0" at the front of $x$. Thus, it will not affect the integer value of $n=\sti{x}$.

@@

Numeric PFA with these restrictions would satisfy our primary objective, that is, they would induce linear formulae and would ``easily" and completely cover all numerals.
%
A last problem still needs to be solved before they can be efficient in practice. 
%
Recall that the character variables can be assigned $\epsilon$.  
%
Therefore, a single chain of $k$ interesting digits, $k\leq m$, can be by $A^m$ represented in $k\choose m$ ways, each corresponding to one possible interleaving of $k$ digits with $m-k$ epsilons. 
%
%\lh{I am not sure whether it is really impossible to make a small formula}
%
This may  lead to a formula of   exponential size  when defining  the value of $n$.
%
In order to eliminate this potential blow-up in the size of the formula, we add to $\paf^m$ an additional constraint that forces all epsilons to be \emph{shifted} behind the least significant digit. This will leave us with only one interleaving. This is the formula
$$
\Psi_{\mathsf{shift}}^m \defeq \bigwedge_{1\leq i \leq m} v_i \neq \noepsilon \implies v_{i-1}\neq \noepsilon \ .
$$
Last, since this restriction is meaningful only when the string is indeed a numeral,  
we also define the constraint representing the strings which are not numerals, the formula
$$
\Psi_{\mathsf{NaN}}^m \defeq  \bigvee_{i\in [1,m]} v_i>9
$$ 
and define the final form of the interpretation restriction used by $A^m$ as 
$$
\paf^m \defeq  \Psi_{\mathsf{NaN}}^m \lor (\Psi_{v_0} \land \Psi_{\mathsf{shift}}^m)\ .
$$
Consequently, we design our domain restrictions $\dr$ so that for string variables $x$ that appear within string-integer constraints,  $\dr(x)$ is a numberic PFA $(A^m,\paf^m),m\in \nat$.
